---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

We first read the raw data from the `.zip` file that (we assume) is in our current working directory into the data frame `data`.

```{r}
if (!file.exists("activity.csv")){
  unzip("activity.zip")
}

data <- read.csv("activity.csv")
str(data)
```

We convert the date column into `POSIXct` format, which will be useful later when applying the ``weekdays()` function. Although irrelevant, we set the time zone to GMT. 

```{r}
data$date <- as.POSIXct(data$date, tz="GMT")
str(data)
```


## What is mean total number of steps taken per day?

We make a data frame containing the total number of steps for each day. We ignore `NA` values.

```{r}
totalstepsPerDay <- aggregate(data$steps, by = list(data$date), FUN=sum)
names(totalstepsPerDay) <- c("date", "totalsteps")
head(totalstepsPerDay)
```

We make a histogram of the total number of steps taken each day.

```{r}
hist(totalstepsPerDay$totalsteps, breaks=33, main="Histogram of total number of steps per day", xlab="Total steps per Day", col="darkblue")
```

Using the total number of steps taken each day, we calculate the mean...

```{r}
mean(totalstepsPerDay$totalsteps, na.rm=TRUE)
```

...and the median of these numbers.

```{r}
median(totalstepsPerDay$totalsteps, na.rm=TRUE)
```


## What is the average daily activity pattern?

We make a data frame containing the average number of steps taken in every 5-minute interval, averaged across all days and removing NA values. 

```{r}
averageStepsPerInterval <- aggregate(data$steps, by = list(data$interval), FUN=mean, na.rm=TRUE)
names(averageStepsPerInterval) <- c("interval", "averagesteps")
head(averageStepsPerInterval)
```

We make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.

```{r}
plot(seq_along(averageStepsPerInterval$interval), averageStepsPerInterval$averagesteps, type="l", main="Average number of steps per 5-minute interval", xlab="5-minute interval (1st interval is [00:00-00:05])", ylab="Average number of steps")
```

We can calculate which interval has the maximum number of steps on average accross all days.  

```{r}
which.max(averageStepsPerInterval$averagesteps)
```

We can also give the interval in a more proper format.

```{r}
leftIndex<-which.max(averageStepsPerInterval$averagesteps)
rightIndex<-(leftIndex+1) %% dim(averageStepsPerInterval)[1] 
maxInterval <- paste("[", floor(averageStepsPerInterval$interval[leftIndex]/100), ":", averageStepsPerInterval$interval[leftIndex]%%100, "-", floor(averageStepsPerInterval$interval[rightIndex]/100), ":", averageStepsPerInterval$interval[rightIndex]%%100, "]", sep="")
```
Therefore, the 5-minute interval with the maximum number of steps on average accross all days is `r maxInterval`.


## Imputing missing values

We calculate the total number of rows having `NA`'s in the number of steps.

```{r}
sum(is.na(data$steps))
```

We replace each `NA` value with the average number of steps in the specific 5-minute interval, accross all days, thus creating the data frame `dataNoNA`. Of course, the steps are automatically coerced into numeric (instead of integer).

```{r}
dataNoNA <- data
dataNoNA$steps[is.na(data$steps)] <- rep(averageStepsPerInterval$averagesteps, times = dim(data)[1]/(24*60/5))[is.na(data$steps)]
head(dataNoNA)
```

We perform similar analysis. In particular, we make a data frame containing the total number of steps for each day using `dataNoNA` this time.

```{r}
totalstepsPerDayNoNA <- aggregate(dataNoNA$steps, by = list(dataNoNA$date), FUN=sum)
names(totalstepsPerDayNoNA) <- c("date", "totalsteps")
head(totalstepsPerDayNoNA)
```

We make a histogram of the total number of steps taken each day.

```{r}
hist(totalstepsPerDayNoNA$totalsteps, breaks=33, main="Histogram of total number of steps per day without NAs", xlab="Total steps per Day", col="darkgreen")
```

We can see that this histogram differs from the corresponding histogram for the original dataframe `data` in the y-axis, since there are a lot of `NA` values taken into account now. 


Using the total number of steps taken each day, we calculate the mean...

```{r}
mean(totalstepsPerDayNoNA$totalsteps, na.rm=TRUE)
```

...and the median of these numbers.

```{r}
median(totalstepsPerDayNoNA$totalsteps, na.rm=TRUE)
```

We can now see that the values for the mean in the processed data frame `dataNoNA` is the same as in the original data frame `data`, and this is because we chose to replace `NA` values with the mean. On the other hand, the median is different from the case where `NA`'s where not taken into account.


## Are there differences in activity patterns between weekdays and weekends?

We create a new factor variable `weekday`, indicating whether a given date is a weekday or weekend day. We use the `dataNoNA` data frame.

```{r}
typeOfDay <-function(date){
  if (weekdays(date) %in% c("Saturday", "Sunday")){
    return("Weekend")
  } else {
      return("Weekday")
  }
} 

dataNoNA$weekday <- sapply(dataNoNA$date, typeOfDay)
dataNoNA$weekday <- as.factor(dataNoNA$weekday)
str(dataNoNA)
```

We now find the mean number of steps accross all weekdays and accross all weekends.

```{r}
averageStepsPerIntervalAndWeekdayNoNA <- aggregate(data$steps, by = list(dataNoNA$interval, dataNoNA$weekday), FUN=mean, na.rm=TRUE)
names(averageStepsPerIntervalAndWeekdayNoNA) <- c("interval", "weekday", "averagesteps")
head(averageStepsPerIntervalAndWeekdayNoNA)
```

We make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
library(lattice)
xyplot(averageStepsPerIntervalAndWeekdayNoNA$averagesteps ~ averageStepsPerIntervalAndWeekdayNoNA$interval | averageStepsPerIntervalAndWeekdayNoNA$weekday, type="l", ylab="Average number of steps", xlab="5-minute interval \n (1st interval is [00:00-00:05])", main="Average number of steps \n per 5-minute interval", layout(c(1,2,1)))
```